---
output: rmarkdown::github_document
---

# fable.gam

This package provides a tidy R interface to time-series modelling using [generalised additive models](https://en.wikipedia.org/wiki/Generalized_additive_model) (GAMs) using [fable](https://github.com/tidyverts/fable). This package makes use of the [mgcv package](https://cran.r-project.org/package=mgcv) for R. While not particularly common, GAMs have shown incredible utility in time-series modelling, such as in the context of [palaeoecological data](https://www.frontiersin.org/journals/ecology-and-evolution/articles/10.3389/fevo.2018.00149/full). This package aims to incorporate the broad conceptual approach of using [structural time-series models](https://blog.tensorflow.org/2019/03/structural-time-series-modeling-in.html) (i.e., decomposing a time series into its trend, seasonality, error, and other components and modelling them additively) within a GAM setup into the incredible `fable` forecasting framework.

```{r, include = FALSE}
knitr::opts_chunk$set(
comment = NA, fig.width = 12, fig.height = 8, cache = FALSE)
```

## Installation

You can install the development version of `fable.gam` from GitHub using the following:

```{r eval = FALSE}
devtools::install_github("hendersontrent/fable.gam")
```

## Quick tour

```{r, warning=FALSE, message=FALSE}
library(tsibble)
library(dplyr)
library(ggplot2)
library(fable)
library(fable.gam)
library(tsibbledata)
```

Just like in the [`fable` vignette](https://fable.tidyverts.org/articles/fable.html) we are going to try and forecast the number of domestic travellers to Melbourne, Australia. In the [`tsibble::tourism`](https://tsibble.tidyverts.org/reference/tourism.html) data set, this can be further broken down into 4 reasons of travel: `“business”`, `“holiday”`, `“visiting friends and relatives”` and `“other reasons”`. The variable we are going to try and forecast is the number of overnight trips (000s) represented by the `Trips` variable.

Thanks to the excellent `tsibble` data structure for storing time-series data in R we know that this data is sampled quarterly. With that in mind, we are going to fit a simple time-series GAM using a (non)linear trend term and a seasonal term with a periodicity of 4. This is made easy in `fable.gam` through the usage of the `fabletools` 'special' functions `trend` and `season` which have been modified in `fable.gam` for the purposes of GAMs. Under the hood, `fable.gam` parses a `trend()` call as modelling a smooth function over time to capture temporal effects and a `season()` call as a smooth function using a [cyclic cubic basis spline](https://fromthebottomoftheheap.net/2014/05/09/modelling-seasonal-data-with-gam/) to ensure that over the duration of the time series, the start and end of a seasonal term connects (e.g., for data measured on a monthly basis, the end of final month -- December -- needs to connect continuously to the start of the following January).

Here is how easy this is to do in `fable.gam` through the `GAM()` function integrated into the `fable` interface:

```{r}
tourism_melb <- tsibble::tourism |>
  filter(Region == "Melbourne")

fit <- tourism_melb |>
  model(mygam = GAM(Trips ~ trend() + season(4)))
```

We can then easily pipe into the rest of the `fable` functionality, such as the `forecast` to produce forecasts. Here are automatic forecasts using the GAM for the next 5 years:

```{r}
fc <- fit |>
  forecast(h = "5 years")

fc |>
  autoplot(tourism_melb)
```

We could then quantify model accuracy using the `accuracy` function in `fable`:

```{r}
fit |>
  accuracy() |>
  arrange(MASE)
```

Another common task is to extract point forecasts and confidence intervals from the forecast distribution. The [`hilo`](https://pkg.mitchelloharawild.com/distributional/reference/hilo.html) function from the [`distributional`](https://github.com/mitchelloharawild/distributional) package knows how to automatically handle models fit in `fable`:

```{r}
fc |>
  hilo(level = c(80, 95))
```

Hopefully this is starting to highlight the power of why integrating new methods into the `fable` framework rather than writing bespoke pipelines is so powerful!

### Comparison to other common time-series methods

We can perform a quick sense-check of the approach against more commonly-used forecasting methods such as [exponential smoothing](https://otexts.com/fpp3/expsmooth.html). Here we will just specify an additive trend for the ETS model and let the other components be determined automatically. We will also log-transform the response variable to show that `fable.gam` handles these transformations automatically as well, just like all models in `fable`.

```{r}
tourism_melb |>
  model(
    mygam = GAM(log(Trips) ~ trend() + season(4)),
    ets = ETS(log(Trips) ~ trend("A"))
  ) |>
  forecast(h = "5 years") |>
  autoplot(tourism_melb)
```

### Controlling the smooth function of the time trend

In addition to the `trend()` and `season()` special functions, `fable.gam` also includes the special `trend2()` which lets users control the smooth term fit to the temporal trend. `trend2()` takes arguments `k` (the dimension of the basis used to represent the smooth term; i.e., the number of knots) and `bs` (a two letter character string indicating the smoothing basis to use). Here is an example where we specify a basis of 5 and a [Gaussian process](https://en.wikipedia.org/wiki/Gaussian_process) smooth:

```{r}
tourism_melb |>
  model(mygam = GAM(Trips ~ trend2(k = 5, bs = "gp") + season(4))) |>
  forecast(h = "5 years") |>
  autoplot(tourism_melb)
```

### Model fit

The suite of `fable` functions for interpreting trained models are also available to `GAM` models fit using `fable.gam`. `glance()` is one such example:

```{r}
tourism_melb |>
  model(mygam = GAM(Trips ~ trend2(k = 5, bs = "gp") + season(4))) |>
  glance()
```

## A more complicated example

`fable.gam` also permits the ability to specify autocorrelated errors to capture residual serial correlation, just like you can do using a `gamm` instead of a `gam` in `mgcv`. Here is another example from [`fpp3`](https://otexts.com/fpp3/seasonal-arima.html) which is seasonal, autocorrelated data with a small trend (note the addition of the `fable.gam` special `errors(ar = 2)` which captures the autocorrelation at lag 2 identified in `fpp3`).

```{r, message=FALSE, warning=FALSE}
h02 <- tsibbledata::PBS |>
  filter(ATC2 == "H02") |>
  summarise(Cost = sum(Cost) / 1e6)

h02 |>
  model(
    mygam = GAM(log(Cost) ~ trend() + season(12) + errors(ar = 2)),
    arima = ARIMA(log(Cost) ~ 0 + pdq(3, 0, 1) + PDQ(0, 1, 2))
    ) |>
  forecast() |>
  autoplot(h02) +
  labs(y = " $AU (millions)",
       title = "Corticosteroid drug scripts (H02) sales") +
  theme(legend.position = "bottom")
```

## Development notes

`fable.gam` is very much a work in progress. Not all current `fable` model functionality has been integrated yet and there may be issues with model estimation until further testing and validation is performed.
