---
title: "Introduction to fable.gam"
author: "Trent Henderson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Introduction to fable.gam}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 5,
  fig.width = 6,
  warning = FALSE,
  fig.align = "center"
)
```

```{r setup, message = FALSE, warning = FALSE}
library(tsibble)
library(dplyr)
library(fable)
library(fable.gam)
```

This package provides a tidy R interface to time-series modelling using [generalised additive models](https://en.wikipedia.org/wiki/Generalized_additive_model) (GAMs) using [fable](https://github.com/tidyverts/fable). This package makes use of the [mgcv package](https://cran.r-project.org/package=mgcv) for R. While not particularly common, GAMs have shown incredible utility in time-series modelling, such as in the context of [palaeoecological data](https://www.frontiersin.org/journals/ecology-and-evolution/articles/10.3389/fevo.2018.00149/full). This package aims to incorporate the broad conceptual approach of using [structural time-series models](https://blog.tensorflow.org/2019/03/structural-time-series-modeling-in.html) within a GAM setup into the incredible `fable` forecasting framework.

```{r, include = FALSE}
knitr::opts_chunk$set(
comment = NA, fig.width = 12, fig.height = 8, cache = FALSE)
```

## Walkthrough of package functionality

Just like in the [`fable` vignette](https://fable.tidyverts.org/articles/fable.html) we are going to try and forecast the number of domestic travellers to Melbourne, Australia. In the [`tsibble::tourism`](https://tsibble.tidyverts.org/reference/tourism.html) data set, this can be further broken down into 4 reasons of travel: `“business”`, `“holiday”`, `“visiting friends and relatives”` and `“other reasons”`. The variable we are going to try and forecast is the number of overnight trips (000s) represented by the `Trips` variable. We can load the dataset and visualise the time series as follows:

```{r}
tourism_melb <- tourism %>%
  filter(Region == "Melbourne")

tourism_melb %>%
  autoplot(Trips)
```

Thanks to the excellent `tsibble` data structure for storing time-series data in R we know that this data is sampled quarterly. With that in mind, we are going to fit a simple time-series GAM using a (non)linear trend term and a seasonal term with a periodicity of 4. This is made easy in `fable.gam` through the usage of the `fabletools` 'special' functions `trend` and `season` which have been modified in `fable.gam` for the purposes of GAMs. Under the hood, `fable.gam` parses a `trend()` call as modelling a smooth function over time to capture temporal effects and a `season()` call as a smooth function using a [cyclic cubic basis spline](https://fromthebottomoftheheap.net/2014/05/09/modelling-seasonal-data-with-gam/) to ensure that over the duration of the time series, the start and end of a seasonal term connects (e.g., for data measured on a monthly basis, the end of final month -- December -- needs to connect continuously to the start of the following January).

Here is how easy this is to do in `fable.gam` through the `GAM()` function integrated into the `fable` interface:

```{r}
fit <- tourism_melb %>%
  model(mygam = GAM(Trips ~ trend() + season(4)))
```

We can then easily pipe into the rest of the `fable` functionality, such as the `forecast` to produce forecasts. Here are automatic forecasts using the GAM for the next 5 years:

```{r}
fc <- fit %>%
  forecast(h = "5 years")

fc %>%
  autoplot(tourism_melb)
```

We could then quantify model accuracy using the `accuracy` function in `fable`:

```{r}
fit %>%
  accuracy() %>%
  arrange(MASE)
```

Another common task is to extract point forecasts and confidence intervals from the forecast distribution. The [`hilo`](https://pkg.mitchelloharawild.com/distributional/reference/hilo.html) function from the [`distributional`](https://github.com/mitchelloharawild/distributional) package knows how to automatically handle models fit in `fable`:

```{r}
fc %>%
  hilo(level = c(80, 95))
```

Hopefully this is starting to highlight the power of why integrating new methods into the `fable` framework rather than writing bespoke pipelines is so powerful!

### Comparison to other common time-series methods

We can perform a quick sense-check of the approach against more commonly-used forecasting methods such as [exponential smoothing](https://otexts.com/fpp3/expsmooth.html). Here we will just specify an additive trend for the ETS model and let the other components be determined automatically:

```{r}
tourism_melb %>%
  model(
    mygam = GAM(Trips ~ trend() + season(4)),
    ets = ETS(Trips ~ trend("A"))
  ) %>%
  forecast(h = "5 years") %>%
  autoplot(tourism_melb)
```

Or maybe we want to compare against an [ARIMA model](https://otexts.com/fpp3/arima.html):

```{r}
tourism_melb %>%
  model(
    mygam = GAM(Trips ~ trend() + season(4)),
    arima = ARIMA(Trips)
  ) %>%
  forecast(h = "5 years") %>%
  autoplot(tourism_melb)
```
